---
title: "GO Analysis"
author: "Ha Tran"
date: "08/09/2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      fig.width = 11)
```

# Data Setup

### Load Library
```{r load libraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)

# Visualisation:
library(kableExtra)
library(ggplot2)
library(grid)
library(pander)
library(cowplot)

# Custom ggplot 
library(ggbiplot)
library(ggrepel)

# Bioconductor packages:
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(org.Hs.eg.db)
# library(enrichplot)
library(biomaRt)

```


### Import RDS Data

DGElist object containing the raw feature count, sample metadata, and gene metadata, created in the Set Up stage. 

```{r importData}
# load DGElist previously created in the set up
dge <- readRDS(here::here("0_data/rds_object/dge.rds"))
designMatrix <- readRDS(here::here("0_data/rds_object/designMatrix.rds"))
contrast <- readRDS(here::here("0_data/rds_object/contrastMatrix.rds"))

treat <- readRDS(here::here("0_data/rds_object/treat.rds"))
treat_filtered <- readRDS(here::here("0_data/rds_object/treat_filtered.rds"))

# qlf <- readRDS(here::here("0_data/rds_object/qlf.rds"))
# qlf_filtered <- readRDS(here::here("0_data/rds_object/qlf_filtered.rds"))
```

# Gene Ontology (GO) Enrichment Analysis

### Using the `clusterProfiler::enrichGO` package {.tabset}

GO enrichment analysis is performed using the DGE obtained from GLM:QLF + TREAT.

The top 30 most significant GO terms are displayed in relation to their respective comparison

```{r treat_enrichGO}
comparison_group=colnames(contrast) %>% as.data.frame()
goSummaries <- url("https://uofabioinformaticshub.github.io/summaries2GO/data/goSummaries.RDS") %>%
  readRDS() 
minPath <- 3

# Create list object
enrichGO_treat=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the enrichGo list with all GO terms
  GOresults <- enrichGO(gene = treat_filtered[[x]]$entrezid,
                        keyType = "ENTREZID",
                        ont = "ALL",
                        OrgDb = org.Hs.eg.db,
                        pAdjustMethod = "fdr")

  enrichGO_treat[[x]] <- GOresults@result
  
  enrichGO_treat[[x]] <- enrichGO_treat[[x]] %>% rownames_to_column("id") %>%
  left_join(goSummaries) %>%
  dplyr::filter(shortest_path >= minPath) %>%
  column_to_rownames("id")

  #filter only significant GO terms
  # enrichGO_treat_sig[[x]] <- enrichGO_treat[[x]] %>% filter(pvalue <= 0.05)
  
}

# save rds object for use in downstream GO and KEGG analysis
enrichGO_treat %>% saveRDS(file = here::here("0_data/rds_object/enrichGO_treat.rds"))
# enrichGO_treat_sig %>% saveRDS(file = here::here("0_data/rds_object/enrichGO_treat_filtered.rds"))

#save each unfiltered comparison group in the output directory
writexl::write_xlsx(x = enrichGO_treat,
                    path = here::here("3_output/enrichGO_treat.xlsx"))
  

# GOnetwork_plot <- enrichGO_treat[["CKI-Omt vs CKI"]] %>% as.data.frame() %>%
#   cnetplot(cex_label_gene = 0.4,
#            showCategory = nrow(enrichGO_treat))
# GOnetwork_plot

```

#### Untreated vs CKI

```{r}
enrichGO_treat[["UT vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID", "shortest_path", "longest_path", "terminal_node", "ontology")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Mac vs CKI

```{r}
enrichGO_treat[["CKI-Mac vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID", "shortest_path", "longest_path", "terminal_node", "ontology")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Nme vs CKI

```{r}
enrichGO_treat[["CKI-Nme vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID", "shortest_path", "longest_path", "terminal_node", "ontology")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Omt vs CKI

```{r}
enrichGO_treat[["CKI-Omt vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID", "shortest_path", "longest_path", "terminal_node", "ontology")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Tri vs CKI

```{r}
enrichGO_treat[["CKI-Tri vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID", "shortest_path", "longest_path", "terminal_node", "ontology")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```




