---
title: "GO Analysis"
author: "Ha Tran"
date: "08/09/2021"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  fig.width = 11
)
```

# Data Setup

### Load Library
```{r load libraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)

# Visualisation:
library(kableExtra)
library(ggplot2)
library(grid)
library(pander)
library(cowplot)

# Custom ggplot
library(ggbiplot)
library(ggrepel)

# Bioconductor packages:
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(org.Hs.eg.db)
# library(enrichplot)
library(biomaRt)

theme_set(theme_minimal())
```


### Import RDS Data

DGElist object containing the raw feature count, sample metadata, and gene metadata, created in the Set Up stage. 

```{r importData}
# load DGElist previously created in the set up
dge <- readRDS(here::here("0_data/rds_object/dge.rds"))
designMatrix <- readRDS(here::here("0_data/rds_object/designMatrix.rds"))
contrast <- readRDS(here::here("0_data/rds_object/contrastMatrix.rds"))
treat <- readRDS(here::here("0_data/rds_object/treat.rds"))
treat_sig <- readRDS(here::here("0_data/rds_object/treat_sig.rds"))
```

# Gene Ontology (GO) Enrichment Analysis
Gene ontology can be perfomred with many packages. `edgeR` can be used to perform GO enrihcment analysis 

### Using the `edgeR::goanna` package {.tabset}

```{r treat_goana}
comparison_group <- colnames(contrast) %>% as.data.frame()
goSummaries <- url("https://uofabioinformaticshub.github.io/summaries2GO/data/goSummaries.RDS") %>%
  readRDS()
minPath <- 3

# Create list object
goana_treat <- list()
goana_treat_sig <- list()

for (i in 1:ncol(contrast)) {
  # at each iteration, let x = name of each contrast group
  x <- comparison_group[i, ]

  # populate the goanna list with all GO terms
  goana_treat[[x]] <- edgeR::goana.DGELRT(
    de = treat[[x]],
    geneid = treat[[x]]$genes$entrezid,
    FDR = 0.05,
    species = "Hs",
    trend = T,
    ont = "ALL"
  )

  # extract the top 300 enriched results
  goana_treat_sig[[x]] <- topGO(results = goana_treat[[x]], ontology = c("MF", "BP", "CC"), number = 300)

  # bind with goSummary to filter out ontologies above level 3
  goana_treat_sig[[x]] <- goana_treat_sig[[x]] %>%
    rownames_to_column("id") %>%
    left_join(goSummaries) %>%
    dplyr::filter(shortest_path >= minPath) %>%
    column_to_rownames("id")
}
```

<!-- ### Using the `clusterProfiler::enrichGO` package {.tabset} -->

<!-- GO enrichment analysis is performed using the DGE obtained from GLM:QLF + TREAT. -->

<!-- The top 30 most significant GO terms are displayed in relation to their respective comparison -->

```{r treat_enrichGO, eval=FALSE}
comparison_group <- colnames(contrast) %>% as.data.frame()
goSummaries <- url("https://uofabioinformaticshub.github.io/summaries2GO/data/goSummaries.RDS") %>%
  readRDS()
minPath <- 3

# Create list object
enrichGO_treat <- list()
for (i in 1:ncol(contrast)) {
  # at each iteration, let x = name of each contrast group
  x <- comparison_group[i, ]

  # populate the enrichGo list with all GO terms
  GOresults <- enrichGO(
    gene = treat_filtered[[x]]$entrezid,
    keyType = "ENTREZID",
    ont = "ALL",
    OrgDb = org.Hs.eg.db,
    pAdjustMethod = "fdr"
  )

  enrichGO_treat[[x]] <- GOresults@result

  enrichGO_treat[[x]] <- enrichGO_treat[[x]] %>%
    rownames_to_column("id") %>%
    left_join(goSummaries) %>%
    dplyr::filter(shortest_path >= minPath) %>%
    column_to_rownames("id")
}
```

#### Untreated vs CKI

```{r}
goana_treat_sig[["UT vs CKI"]][1:30, ] %>%
  as.data.frame() %>%
  dplyr::select(-c("shortest_path", "longest_path", "terminal_node", "ontology")) %>%
  kable(digits = Inf) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

#### CKI-Mac vs CKI

```{r}
goana_treat_sig[["CKI-Mac vs CKI"]][1:30, ] %>%
  as.data.frame() %>%
  dplyr::select(-c("shortest_path", "longest_path", "terminal_node", "ontology")) %>%
  kable(digits = Inf) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

#### CKI-Nme vs CKI

```{r}
goana_treat_sig[["CKI-Nme vs CKI"]][1:30, ] %>%
  as.data.frame() %>%
  dplyr::select(-c("shortest_path", "longest_path", "terminal_node", "ontology")) %>%
  kable(digits = Inf) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

#### CKI-Omt vs CKI

```{r}
goana_treat_sig[["CKI-Omt vs CKI"]][1:30, ] %>%
  as.data.frame() %>%
  dplyr::select(-c("shortest_path", "longest_path", "terminal_node", "ontology")) %>%
  kable(digits = Inf) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

#### CKI-Tri vs CKI

```{r}
goana_treat_sig[["CKI-Tri vs CKI"]][1:30, ] %>%
  as.data.frame() %>%
  dplyr::select(-c("shortest_path", "longest_path", "terminal_node", "ontology")) %>%
  kable(digits = Inf) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "600px")
```

### Enriched GO terms
```{r dotPlot, fig.width=8, fig.height=9}
top_go <- list()
for (i in 1:ncol(contrast)) {
  # at each iteration, let x = name of each contrast group
  x <- comparison_group[i, ]

  top_go[[x]] <- goana_treat_sig[[x]][1:7,]
}
# combine all list from big merged_go into one
merged_go <- as.data.frame(do.call(rbind, lapply(top_go, as.data.frame))) %>% rownames_to_column("group")

# remove KEGG ID from group names
merged_go$group <- gsub(pattern = "\\..*", "", merged_go$group)

# direction column
merged_go <- merged_go %>% dplyr::mutate(direction = case_when
(
  Up > Down ~ "Up-\nregulated\n",
  Up < Down ~ "Down-\nregulated\n",
  TRUE ~ "Mutual"
))

# geneRatio column 
merged_go <- merged_go %>% dplyr::mutate(geneRatio = case_when
(
  Up > Down & group == "UT vs CKI" ~ Up / 215,
  Up > Down & group == "CKI-Omt vs CKI" ~ Up / 689,
  Up > Down & group == "CKI-Mac vs CKI" ~ Up / 353,
  Up > Down & group == "CKI-Nme vs CKI" ~ Up / 382,
  Up > Down & group == "CKI-Tri vs CKI" ~ Up / 358,
  Up < Down & group == "UT vs CKI" ~ Down / 215,
  Up < Down & group == "CKI-Omt vs CKI" ~ Down / 689,
  Up < Down & group == "CKI-Mac vs CKI" ~ Down / 353,
  Up < Down & group == "CKI-Nme vs CKI" ~ Down / 382,
  Up < Down & group == "CKI-Tri vs CKI" ~ Down / 358,
))

# pvalue column
merged_go <- merged_go %>% dplyr::mutate(pValue = case_when
(
  Up > Down ~ P.Up,
  Up < Down ~ P.Down,
  TRUE ~ P.Up
))

merged_go <- merged_go %>% filter(pValue <= 0.05)

# Transform the column 'Description' into factors
merged_go$Term <- as.factor(merged_go$Term)

# Transform FDR values by -log10('FDR values')
merged_go$"|log10(FDR)|" <- -(log10(merged_go$pValue))

# Change factor order
merged_go$group <- factor(merged_go$group, levels = c("UT vs CKI", "CKI-Omt vs CKI", "CKI-Mac vs CKI", "CKI-Nme vs CKI", "CKI-Tri vs CKI"))

# merged_go$Term <- factor(merged_go, levels = rev(levels(merged_go$Term)))

group.labs <- merged_go$group

ggplot(merged_go, aes(x = Term, y = group)) +
  geom_point(data = merged_go, aes(x = Term, y = group, size = geneRatio, colour = `|log10(FDR)|`, shape = direction), alpha = .9) +
  # scale_y_discrete(labels =group.labs)+
  scale_shape_manual(values = c(15, 16, 17)) +
  scale_color_gradient(low = "dodgerblue3", high = "firebrick3", limits = c(0, NA)) +
  coord_flip() +
  theme_bw() +
  theme(
    axis.ticks.length = unit(-0.1, "cm"),
    axis.text.x = element_text(margin = margin(5, 5, 0, 5, "pt"), angle = 45, hjust = 1),
    axis.text.y = element_text(margin = margin(5, 5, 5, 5, "pt")),
    axis.text = element_text(color = "black"),
    axis.title.x = element_blank()
  ) +
  xlab(label = "") +
  labs(color = expression("-log"[10] * "FDR"), size = "Gene Ratio", shape = "Expression")

ggsave(filename = "dot_goanna_treat.svg", plot = last_plot(), path = here::here("2_plots/3_go/"))
```


# Export Data
```{r export}
writexl::write_xlsx(x = goana_treat, path = here::here("3_output/goana_treat.xlsx"))
# writexl::write_xlsx(x = enrichGO_treat, path = here::here("3_output/enrichGO_treat.xlsx"))
```
