---
title: "KEGG Analysis"
author: "Ha Tran"
date: "19/10/2021"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      fig.width = 11)
```

# Data Setup

### Load Library
```{r load libraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)
library(KEGGREST)

# Visualisation:
library(kableExtra)
library(ggplot2)
library(grid)
library(pander)
library(cowplot)
library(pheatmap)

# Custom ggplot 
library(ggbiplot)
library(ggrepel)

# Bioconductor packages:
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(biomaRt)

# Set ggplot theme
theme_set(theme_minimal())
```


### Import RDS Data

DGElist object containing the raw feature count, sample metadata, and gene metadata, created in the Set Up stage. 

```{r importData}
# load DGElist previously created in the set up
dge <- readRDS(here::here("0_data/rds_object/dge.rds"))
designMatrix <- readRDS(here::here("0_data/rds_object/designMatrix.rds"))
contrast <- readRDS(here::here("0_data/rds_object/contrastMatrix.rds"))
treat <- readRDS(here::here("0_data/rds_object/treat.rds"))
treat_all <- readRDS(here::here("0_data/rds_object/treat_all.rds"))
treat_sig <- readRDS(here::here("0_data/rds_object/treat_sig.rds"))
```

# KEGG enrichment analysis

### Using the `clusterProfiler::enrichKEGG` package {.tabset}

KEGG pathway enrichment analysis is performed with GLM:QLF + TREAT.

The top 30 most significant KEGG terms are displayed in relation to their respective comparison. KEGG terms with `qvalue <= 0.2` are considered as significant

```{r treatKEGGanalysis}
comparison_group=colnames(contrast) %>% as.data.frame()
# Create list object
enrichKEGG_treat=list()
enrichKEGG_treat_sig=list()

for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the enrichGo list with all KEGG terms
  KEGGresults <- clusterProfiler::enrichKEGG(gene = treat_sig[[x]]$entrezid,
                        organism = "hsa",
                        pvalueCutoff = 0.05,
                        pAdjustMethod = "fdr") 
  
  enrichKEGG_treat[[x]] <- KEGGresults@result %>% as.data.frame()
 
  #filter only significant KEGG terms
  enrichKEGG_treat_sig[[x]] <- enrichKEGG_treat[[x]] %>% dplyr::filter(qvalue <= 0.2)
} 

```

### `edgeR::kegga` package {.tabset}

Through edgeR, the proportion of upregulation and downregulated genes enriching the KEGG pathway can be determined 
```{r kegga}
comparison_group=colnames(contrast) %>% as.data.frame()
# Create list object
kegga_treat=list()
kegga_treat_sig=list()
# kegga_treat_sig=list()

for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the enrichGo list with all KEGG terms
  kegga_treat[[x]] <- edgeR::kegga.DGELRT(de = treat[[x]],
                                    geneid = treat[[x]]$genes$entrezid,
                                    species = "Hs",
                                    FDR = 0.05, 
                                    trend = F) 
  
  kegga_treat_sig[[x]] <- topKEGG(results = kegga_treat[[x]], number = Inf)
}
```

#### UT vs CKI

```{r}
kegga_treat_sig[["UT vs CKI"]][1:30,] %>% as.data.frame() %>% kable(digits = Inf) %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Mac vs CKI

```{r}
kegga_treat_sig[["CKI-Mac vs CKI"]][1:30,] %>% as.data.frame() %>% kable(digits = Inf) %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Nme vs CKI

```{r}
kegga_treat_sig[["CKI-Nme vs CKI"]][1:30,] %>% as.data.frame() %>% kable(digits = Inf) %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Omt vs CKI

```{r}
kegga_treat_sig[["CKI-Omt vs CKI"]][1:30,] %>% as.data.frame() %>% kable(digits = Inf) %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Tri vs CKI

```{r}
kegga_treat_sig[["CKI-Tri vs CKI"]][1:30,] %>% as.data.frame() %>% kable(digits = Inf ) %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

# Visualisation
```{r dotPlot, fig.width=8, fig.height=9}
top_kegg=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]

top_kegg[[x]] <- kegga_treat_sig[[x]][1:15,]
}
#combine all list from big merged_kegg into one
merged_kegg <- as.data.frame(do.call(rbind, lapply(top_kegg, as.data.frame))) %>% rownames_to_column("group")

#remove KEGG ID from group names
merged_kegg$group <- gsub(pattern = "\\..*", "", merged_kegg$group)

#direction column
merged_kegg <- merged_kegg %>% dplyr::mutate(direction = case_when
                         (Up > Down ~ "Up-\nregulated\n",
                          Up < Down ~ "Down-\nregulated\n",
                          TRUE ~ "Mutual"))

#geneRatio column
merged_kegg <- merged_kegg %>% dplyr::mutate(geneRatio = case_when
                                     (Up > Down & group == "UT vs CKI" ~ Up/215,
                                       Up > Down & group == "CKI-Omt vs CKI" ~ Up/689,
                                        Up > Down & group == "CKI-Mac vs CKI" ~ Up/353,
                                        Up > Down & group == "CKI-Nme vs CKI" ~ Up/382,
                                        Up > Down & group == "CKI-Tri vs CKI" ~ Up/358,
                                      Up < Down & group == "UT vs CKI" ~ Down/215,
                                       Up < Down & group == "CKI-Omt vs CKI" ~ Down/689,
                                        Up < Down & group == "CKI-Mac vs CKI" ~ Down/353,
                                        Up < Down & group == "CKI-Nme vs CKI" ~ Down/382,
                                        Up < Down & group == "CKI-Tri vs CKI" ~ Down/358,
                                       ))

#pvalue colum
merged_kegg <- merged_kegg %>% dplyr::mutate(pValue = case_when
                         (Up > Down ~ P.Up,
                          Up < Down ~ P.Down,
                          TRUE ~ P.Up))

merged_kegg <- merged_kegg %>% filter(pValue <= 0.05)

# Transform the column 'Description' into factors
merged_kegg$Pathway <- as.factor(merged_kegg$Pathway)

# Transform FDR values by -log10('FDR values')
merged_kegg$'|log10(FDR)|' <- -(log10(merged_kegg$pValue))

# Change factor order
merged_kegg$group <- factor(merged_kegg$group, levels = c("UT vs CKI", "CKI-Omt vs CKI", "CKI-Mac vs CKI", "CKI-Nme vs CKI", "CKI-Tri vs CKI"))

merged_kegg$Pathway <- factor(merged_kegg$Pathway, levels = rev(levels(merged_kegg$Pathway)))

group.labs <- merged_kegg$group

ggplot(merged_kegg, aes(x = Pathway, y = group)) +
  geom_point(data=merged_kegg,aes(x=Pathway, y=group, size = geneRatio, colour = `|log10(FDR)|`, shape = direction), alpha=.9)+
  # scale_y_discrete(labels =group.labs)+
  scale_shape_manual(values = c(15,16,17))+
  scale_color_gradient(low = "dodgerblue3", high = "firebrick3", limits=c(0, NA))+
  coord_flip()+
  theme_bw()+
  theme(axis.ticks.length=unit(-0.1, "cm"),
        axis.text.x = element_text(margin=margin(5,5,0,5,"pt"),angle = 45, hjust = 1),
        axis.text.y = element_text(margin=margin(5,5,5,5,"pt")),
        axis.text = element_text(color = "black"),
        axis.title.x=element_blank())+
  xlab(label = "")+
  labs(color=expression("-log"[10]*"FDR"), size="Gene Ratio", shape = "Expression")
  # theme(legend.position = "bottom")
ggsave(filename = "kegga_treat.svg", plot = last_plot(), path = here::here("2_plots/4_kegg/"))
```


```{r pathwayName}
#kegg pathway of interest
kegg_id <- "hsa04110"
kegg_pathway <- KEGGREST::keggGet(kegg_id)
kegg_pathway_name <- kegg_pathway[[1]]$NAME
```

## Heatmap of `r kegg_pathway_name` {.tabset}

### Partial 
The partial heatmap illustrates the logFC of significant DE genes within the `r kegg_pathway_name` pathway. 

```{r partialPathway, ,fig.height=10, fig.width=8}
# kegg_id <- "hsa04657"
# kegg_pathway <- KEGGREST::keggGet(kegg_id)
# kegg_pathway_name <- kegg_pathway[[1]]$NAME
entrezID=list()
matrix_all=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  entrezID[[x]] <- kegg_pathway[[1]]$GENE %>% as.data.frame()
  entrezID[[x]] <- entrezID[[x]][seq(1, nrow(entrezID[[x]]), 2),] %>% as.data.frame()
  colnames(entrezID[[x]]) <- "entrezid"
  
  match <- treat_sig[[x]]$entrezid %in% entrezID[[x]]$entrezid
  matrix_all[[x]] <- treat_sig[[x]][match, c("gene_name", "logFC")]
}

big_matrix <- as.data.frame(do.call(rbind, lapply(matrix_all, as.data.frame))) %>% rownames_to_column("group") 
big_matrix$group <- gsub(pattern = "\\..*", "", big_matrix$group)

big_matrix <- pivot_wider(big_matrix, names_from = group, values_from = logFC) %>% column_to_rownames("gene_name")
# big_matrix[is.na(big_matrix)]=-0.7305
summary(big_matrix)

## this was a waste of time
# huge_matrix <- as.data.frame(do.call(rbind, lapply(treat_all, as.data.frame))) %>% rownames_to_column("group") %>% dplyr::select(c("group", "gene", "logFC"))
# huge_matrix$group <- gsub(pattern = "\\..*", "", huge_matrix$group)
# huge_matrix <- pivot_wider(huge_matrix, names_from = group, values_from = logFC) %>% column_to_rownames("gene")
# huge_matrix <- huge_matrix[rownames(big_matrix),]

  my_palette <- colorRampPalette(c(
  rgb(32,121,226, maxColorValue = 255),
  # rgb(144,203,180, maxColorValue = 255), 
  rgb(255,255,255, maxColorValue = 255), 
  # rgb(251,192,52, maxColorValue = 255), 
  rgb(226,46,45, maxColorValue = 255)))(n = 201)
 
p <- pheatmap(mat = big_matrix,
         cluster_cols = F,
         cluster_rows = F,
         clustering_distance_rows = "euclidean",
         treeheight_row = 30,
         treeheight_col = 20,
         cutree_rows = 4,
         cutree_cols = 3,
         na_col = "white",
         main = paste0("Heatmap of ", kegg_pathway_name,"\n"),
         angle_col =90,
         
         # scale = "row",
         legend_breaks = c(seq(-8, 8, by = 1), max(big_matrix, na.rm = T)),
         legend_labels = c(seq(-8, 8, by = 1), "logCPM\n"),
         # border_color = F,
         border_color = "gray90",
         color = my_palette,
         )

invisible(dev.print(device = svg, here::here("2_plots/4_kegg/heatmap_partial.svg")))
```

```{r partialPathwayOld, eval=FALSE, fig.width=7, fig.height=7}
entrezID=list()
matrix=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  entrezID[[x]] <- enrichKEGG_treat_sig[[x]][,c("ID", "geneID")] %>% as.data.frame()
  entrezID[[x]] <- entrezID[[x]][kegg_id,"geneID"] %>% as.data.frame()
  entrezID[[x]] <- separate_rows(entrezID[[x]],.,sep = "/")
  colnames(entrezID[[x]]) <- "entrezid"
  
  match <- treat_sig[[x]]$entrezid %in% entrezID[[x]]$entrezid
  matrix[[x]] <- treat_sig[[x]][match, c("gene_name", "logFC")]
}

matrix_merged <- 
  matrix$`UT vs CKI` %>% 
  left_join(matrix$`CKI-Mac vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Nme vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Omt vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Tri vs CKI`, by = "gene_name") %>% 
  column_to_rownames("gene_name")

colnames(matrix_merged) <- c("UT\nvs\nCKI", "CKI-Mac\nvs\nCKI", "CKI-Nme\nvs\nCKI", "CKI-Omt\nvs\nCKI", "CKI-Tri\nvs\nCKI")
my_palette <- colorRampPalette(c(
  rgb(32,121,226, maxColorValue = 255),
  # rgb(144,203,180, maxColorValue = 255), 
  rgb(254,248,239, maxColorValue = 255), 
  # rgb(251,192,52, maxColorValue = 255), 
  rgb(226,46,45, maxColorValue = 255)))(n = 201)
pheatmap(mat = matrix_merged,
         clustering_distance_rows = "euclidean",
         treeheight_row = 70,
         legend_labels = "logFC",
         main = paste0("Heatmap of ", kegg_pathway_name),
         # fontsize = 14,
         # fontsize_row = 12,
         # fontsize_number = 10,
         # fontsize_col = 18,
         angle_col =0,
         border_color = "gray70",
         color = my_palette)
invisible(dev.print(svg, here::here("2_plots/4_kegg/heatmap_partial.svg")))
        
```

### Complete Pathway

This heatmap illustrates the logfc of all the genes within the `r kegg_pathway_name`, regardless of their statistical significance. Th

```{r completePathway, fig.height=16, fig.width=8}
entrezID=list()
matrix_all=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  entrezID[[x]] <- kegg_pathway[[1]]$GENE %>% as.data.frame()
  entrezID[[x]] <- entrezID[[x]][seq(1, nrow(entrezID[[x]]), 2),] %>% as.data.frame()
  colnames(entrezID[[x]]) <- "entrezid"
  
  match <- treat_all[[x]]$entrezid %in% entrezID[[x]]$entrezid
  matrix_all[[x]] <- treat_all[[x]][match, c("gene_name", "logFC")]
}
matrix_all_merged <- matrix_all$`UT vs CKI` %>% 
  left_join(matrix_all$`CKI-Mac vs CKI`, by = "gene_name") %>% 
  left_join(matrix_all$`CKI-Nme vs CKI`, by = "gene_name") %>% 
  left_join(matrix_all$`CKI-Omt vs CKI`, by = "gene_name") %>% 
  left_join(matrix_all$`CKI-Tri vs CKI`, by = "gene_name") %>% 
  column_to_rownames("gene_name")
colnames(matrix_all_merged) <- c("UT vs CKI", "CKI-Mac vs CKI", "CKI-Nme vs CKI", "CKI-Omt vs CKI", "CKI-Tri vs CKI")

summary(matrix_all_merged)
p <- pheatmap(mat = matrix_all_merged,
         cluster_cols = T,
         clustering_distance_rows = "euclidean",
         treeheight_row = 60,
         treeheight_col = 20,
         cutree_rows = 4,
         cutree_cols = 3,
         
         main = paste0("Heatmap of ", kegg_pathway_name),
         
         scale = "row",
         legend = T,
         legend_breaks = c(seq(-3, 3, by = .5), 1.7),
         legend_labels = c(seq(-3, 3, by = .5), "Z-Score\n\n"),
         # border_color = F,
         border_color = "gray70",
         color = my_palette,
         )
# ggsave(filename = "heatmap_all.svg",plot = last_plot(), path = here::here("2_plots/4_kegg/"))
invisible(dev.print(width = 8.8, height = 22, device = svg, here::here("2_plots/4_kegg/heatmap_complete.svg")))
         
```

# Export Data
```{r export}
writexl::write_xlsx(x = enrichKEGG_treat, path = here::here("3_output/enrichKEGG_treat_sig.xlsx"))
writexl::write_xlsx(x = kegga_treat, path = here::here("3_output/kegga_treat_sig.xlsx"))
```

