---
title: "KEGG Analysis"
author: "Ha Tran"
date: "19/10/2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      fig.width = 11)
```

# Data Setup

### Load Library
```{r load libraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)
library(KEGGREST)

# Visualisation:
library(kableExtra)
library(ggplot2)
library(grid)
library(pander)
library(cowplot)
library(pheatmap)

# Custom ggplot 
library(ggbiplot)
library(ggrepel)

# Bioconductor packages:
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(biomaRt)

# Set ggplot theme
theme_set(theme_minimal())

theme_update(
  legend.background = element_rect(fill = "transparent", colour = NA),
  legend.box.background = element_rect(fill = "transparent", colour = NA),
  
  panel.background = element_rect(fill = "transparent",colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA),
  
  plot.title = element_text(color = "gray20", size = 24, angle = 0, hjust = 0, vjust = .5, face = "bold"),
  plot.subtitle = element_text(color = "gray25", size = 20, angle = 0, hjust = 0, vjust = .5, face = "plain"),
  legend.title = element_text(color = "gray25", size = 16, angle = 0, hjust = 0, vjust = .5, face = "plain"),
  legend.text = element_text(color = "gray25", size = 12, angle = 0, hjust = 0, vjust = .5, face = "plain"),
  axis.text.x = element_text(color = "gray30", size = 12, angle = 0, hjust = .5, vjust = .5, face = "plain"),
  axis.text.y = element_text(color = "gray30", size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain"),  
  axis.title.x = element_text(color = "gray30", size = 14, angle = 0, hjust = .5, vjust = 0, face = "plain"),
  axis.title.y = element_text(color = "gray30", size = 14, angle = 90, hjust = .5, vjust = .5, face = "plain"))
```


### Import RDS Data

DGElist object containing the raw feature count, sample metadata, and gene metadata, created in the Set Up stage. 

```{r importData}
# load DGElist previously created in the set up
dge <- readRDS(here::here("0_data/rds_object/dge.rds"))
designMatrix <- readRDS(here::here("0_data/rds_object/designMatrix.rds"))
contrast <- readRDS(here::here("0_data/rds_object/contrastMatrix.rds"))

treat <- readRDS(here::here("0_data/rds_object/treat.rds"))
treat_unfiltered <- readRDS(here::here("0_data/rds_object/treat_unfiltered.rds"))
treat_filtered <- readRDS(here::here("0_data/rds_object/treat_filtered.rds"))

```


# KEGG enrichment analysis

### Using the `clusterProfiler::enrichKEGG` package {.tabset}

KEGG pathway enrichment analysis is performed with GLM:QLF + TREAT.

The top 30 most significant KEGG terms are displayed in relation to their respective comparison. KEGG terms with `qvalue <= 0.2` are considered as significant

```{r treatKEGGanalysis}
comparison_group=colnames(contrast) %>% as.data.frame()

# Create list object
enrichKEGG_treat=list()
enrichKEGG_treat_sig=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the enrichGo list with all KEGG terms
  KEGGresults <- clusterProfiler::enrichKEGG(gene = treat_filtered[[x]]$entrezid,
                        organism = "hsa",
                        pvalueCutoff = 0.05,
                        pAdjustMethod = "fdr") 
  
  enrichKEGG_treat[[x]] <- KEGGresults@result %>% as.data.frame()
 
  #filter only significant KEGG terms
  enrichKEGG_treat_sig[[x]] <- enrichKEGG_treat[[x]] %>% dplyr::filter(qvalue <= 0.2)
} 
# save rds object for use in downstream GO and KEGG analysis
# enrichKEGG_treat %>% saveRDS(file = here::here("0_data/rds_object/enrichKEGG_treat_unfiltered.rds"))
enrichKEGG_treat_sig %>% saveRDS(file = here::here("0_data/rds_object/enrichKEGG_treat_filtered.rds"))

#save each unfiltered comparison group in the output directory
writexl::write_xlsx(x = enrichKEGG_treat,
                    path = here::here("3_output/enrichKEGG_treat_unfiltered.xlsx"))

```

#### UT vs CKI

```{r}
enrichKEGG_treat_sig[["UT vs CKI"]][1:12,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Mac vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Mac vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Nme vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Nme vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Omt vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Omt vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Tri vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Tri vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

# Visualisation

```{r pathwayName}
kegg_id <- "hsa04657"
kegg_pathway <- KEGGREST::keggGet(kegg_id)
kegg_pathway_name <- kegg_pathway[[1]]$NAME
```

## Heatmap of `r kegg_pathway_name` {.tabset}

### Partial 

```{r partialPathway}
# kegg_id <- "hsa04657"
# kegg_pathway <- KEGGREST::keggGet(kegg_id)
# kegg_pathway_name <- kegg_pathway[[1]]$NAME

entrezID=list()
matrix=list()

for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  entrezID[[x]] <- enrichKEGG_treat_sig[[x]][,c("ID", "geneID")] %>% as.data.frame()
  entrezID[[x]] <- entrezID[[x]][kegg_id,"geneID"] %>% as.data.frame()
  entrezID[[x]] <- separate_rows(entrezID[[x]],.,sep = "/")
  colnames(entrezID[[x]]) <- "entrezid"
  
  match <- treat_filtered[[x]]$entrezid %in% entrezID[[x]]$entrezid

  matrix[[x]] <- treat_filtered[[x]][match, c("gene_name", "logFC")]
}

matrix_merged <- matrix$`UT vs CKI` %>% 
  left_join(matrix$`CKI-Mac vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Nme vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Omt vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Tri vs CKI`, by = "gene_name") %>% 
  column_to_rownames("gene_name")

colnames(matrix_merged) <- c("UT\nvs\nCKI", "CKI-Mac\nvs\nCKI", "CKI-Nme\nvs\nCKI", "CKI-Omt\nvs\nCKI", "CKI-Tri\nvs\nCKI")


my_palette <- colorRampPalette(c(
  rgb(32,121,226, maxColorValue = 255),
  # rgb(144,203,180, maxColorValue = 255), 
  rgb(254,238,229, maxColorValue = 255), 
  # rgb(251,192,52, maxColorValue = 255), 
  rgb(226,46,45, maxColorValue = 255)))(n = 201)


pheatmap(mat = matrix_merged,
         clustering_distance_rows = "euclidean",
         treeheight_row = 70,
         legend_labels = "logFC",
         main = paste0("Heatmap of ", kegg_pathway_name),
         fontsize = 14,
         fontsize_row = 12,
         fontsize_number = 10,
         fontsize_col = 18,
         angle_col =0,
         border_color = "gray70",
         color = my_palette)

invisible(dev.print(svg, here::here("2_plots/4_kegg/heatmap_partial.svg")))
        
```

### Complete

```{r completePathway, fig.height=15}
# kegg_id <- "hsa04657"
# kegg_pathway <- KEGGREST::keggGet(kegg_id)
# kegg_pathway_name <- kegg_pathway[[1]]$NAME

entrezID=list()
matrix_all=list()

for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  entrezID[[x]] <- kegg_pathway[[1]]$GENE %>% as.data.frame()
  entrezID[[x]] <- entrezID[[x]][seq(1, nrow(entrezID[[x]]), 2),] %>% as.data.frame()

  colnames(entrezID[[x]]) <- "entrezid"
  
  match <- treat_unfiltered[[x]]$entrezid %in% entrezID[[x]]$entrezid

  matrix_all[[x]] <- treat_unfiltered[[x]][match, c("gene_name", "logFC")]
}

matrix_all_merged <- matrix_all$`UT vs CKI` %>% 
  left_join(matrix_all$`CKI-Mac vs CKI`, by = "gene_name") %>% 
  left_join(matrix_all$`CKI-Nme vs CKI`, by = "gene_name") %>% 
  left_join(matrix_all$`CKI-Omt vs CKI`, by = "gene_name") %>% 
  left_join(matrix_all$`CKI-Tri vs CKI`, by = "gene_name") %>% 
  column_to_rownames("gene_name")

colnames(matrix_all_merged) <- c("UT vs CKI", "CKI-Mac vs CKI", "CKI-Nme vs CKI", "CKI-Omt vs CKI", "CKI-Tri vs CKI")

p <- pheatmap(mat = matrix_all_merged,
         cluster_cols = F,
         clustering_distance_rows = "euclidean",
         treeheight_row = 70,
         treeheight_col = 50,
         cutree_rows = 4,
         
         # legend_labels = "logFC",
         main = paste0("Heatmap of ", kegg_pathway_name),
         # cellwidth = 65,
         # cellheight = 25,
         
         fontsize = 14,
         fontsize_row = 12,
         fontsize_number = 10,
         fontsize_col = 18,
         angle_col =90,
         
         scale = "row",
         # legend_breaks = c(seq(-2, 8, by = 2), max(matrix_all_merged)),
         # legend_labels = c(seq(-2, 8, by = 2), "logCPM\n"),
         # border_color = F,
         border_color = "gray70",
         color = my_palette,
         )

# ggsave(filename = "heatmap_all.svg",plot = last_plot(), path = here::here("2_plots/4_kegg/"))
invisible(dev.print(width = 8.5, height = 18, device = svg, here::here("2_plots/4_kegg/heatmap_complete.svg")))
         

```


