---
title: "KEGG Analysis"
author: "Ha Tran"
date: "19/10/2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      fig.width = 11)
```

# Data Setup

### Load Library
```{r load libraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)
library(KEGGREST)

# Visualisation:
library(kableExtra)
library(ggplot2)
library(grid)
library(pander)
library(cowplot)
library(pheatmap)

# Custom ggplot 
library(ggbiplot)
library(ggrepel)

# Bioconductor packages:
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(biomaRt)

# Set ggplot theme
theme_set(theme_bw())

theme_update(legend.background = element_rect(fill = "transparent", colour = NA),
      legend.box.background = element_rect(fill = "transparent", colour = NA),
      
      panel.background = element_rect(fill = "transparent",colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA),
      
      plot.title = element_text(color = "gray20", size = 22, angle = 0, hjust = 0, vjust = .5, face = "bold"),
      axis.text.x = element_text(color = "gray30", size = 12, angle = 0, hjust = .5, vjust = .5, face = "plain"),
      axis.text.y = element_text(color = "gray30", size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain"),  
      axis.title.x = element_text(color = "gray30", size = 12, angle = 0, hjust = .5, vjust = 0, face = "plain"),
      axis.title.y = element_text(color = "gray30", size = 12, angle = 90, hjust = .5, vjust = .5, face = "plain"))
```


### Import RDS Data

DGElist object containing the raw feature count, sample metadata, and gene metadata, created in the Set Up stage. 

```{r importData, eval=FALSE}
# load DGElist previously created in the set up
dge <- readRDS("../0_data/rds_object/dge.rds")
designMatrix <- readRDS("../0_data/rds_object/designMatrix.rds")
contrast <- readRDS("../0_data/rds_object/contrastMatrix.rds")

treat <- readRDS("../0_data/rds_object/treat.rds")
# treat <- readRDS("../0_data/rds_object/treat_unfiltered")
treat_filtered <- readRDS("../0_data/rds_object/treat_filtered.rds")

qlf <- readRDS("../0_data/rds_object/qlf.rds")
qlf_filtered <- readRDS("../0_data/rds_object/qlf_filtered.rds")
```


# KEGG enrichment analysis

### Using the `clusterProfiler::enrichKEGG` package {.tabset}

KEGG pathway enrichment analysis is performed with GLM:QLF + TREAT.

The top 30 most significant KEGG terms are displayed in relation to their respective comparison. KEGG terms with `qvalue <= 0.2` are considered as significant

```{r treatKEGGanalysis}
comparison_group=colnames(contrast) %>% as.data.frame()

# Create list object
enrichKEGG_treat=list()
enrichKEGG_treat_sig=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the enrichGo list with all KEGG terms
  KEGGresults <- clusterProfiler::enrichKEGG(gene = treat_filtered[[x]]$entrezid,
                        organism = "hsa",
                        pvalueCutoff = 0.05,
                        pAdjustMethod = "fdr") 
  
  enrichKEGG_treat[[x]] <- KEGGresults@result %>% as.data.frame()
 
  #filter only significant KEGG terms
  enrichKEGG_treat_sig[[x]] <- enrichKEGG_treat[[x]] %>% dplyr::filter(qvalue <= 0.2)
} 
# save rds object for use in downstream GO and KEGG analysis
# enrichKEGG_treat %>% saveRDS(file = "../0_data/rds_object/enrichKEGG_treat_unfiltered.rds")
enrichKEGG_treat_sig %>% saveRDS(file = "../0_data/rds_object/enrichKEGG_treat_filtered.rds")

#save each unfiltered comparison group in the output directory
writexl::write_xlsx(x = enrichKEGG_treat,
                    path = "../3_output/enrichKEGG_treat_unfiltered.xlsx")

```

#### UT vs CKI

```{r}
enrichKEGG_treat_sig[["UT vs CKI"]][1:12,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Mac vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Mac vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Nme vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Nme vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Omt vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Omt vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

#### CKI-Tri vs CKI

```{r}
enrichKEGG_treat_sig[["CKI-Tri vs CKI"]][1:30,] %>% as.data.frame() %>% dplyr::select(-c("ID")) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover")) %>% scroll_box(height="600px")
```

<!-- #### QLF -->

<!-- not working -->

```{r qlfKEGGanalysis, eval=FALSE}
#why is CKI vs Omt not significant at all?
comparison_group=colnames(contrast) %>% as.data.frame()

# Create list object
enrichKEGG_qlf=list()
enrichKEGG_qlf_sig=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the enrichGo list with all KEGG terms
  KEGGresults <- clusterProfiler::enrichKEGG(gene = qlf_filtered[[x]]$entrezid,
                        organism = "hsa",
                        pAdjustMethod = "fdr") 
  
  enrichKEGG_qlf[[x]] <- KEGGresults@result %>% as.data.frame()
 
  #filter only significant KEGG terms
  enrichKEGG_qlf_sig[[x]] <- enrichKEGG_qlf[[x]] %>% filter(pvalue <= 0.05)
} 
# save rds object for use in downstream GO and KEGG analysis
enrichKEGG_qlf %>% saveRDS(file = "../0_data/rds_object/enrichKEGG_qlf_unfiltered.rds")
enrichKEGG_qlf_sig %>% saveRDS(file = "../0_data/rds_object/enrichKEGG_qlf_filtered.rds")

#save each unfiltered comparison group in the output directory
writexl::write_xlsx(x = enrichKEGG_qlf,
                    path = "../3_output/enrichKEGG_qlf_unfiltered.xlsx")

```

<!-- ### Using the `edgeR::keggaLRT` package -->

<!-- #### TREAT -->

```{r treatKEGGA, eval=FALSE}
# Create list object
kegga_treat=list()
kegga_treat_sig=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the kegga list with all KEGG terms
  kegga_treat[[x]] <- edgeR::kegga.DGELRT(de = treat[[x]], 
                                          geneid = treat[[x]]$genes$entrezid, 
                                          species = "Hs",
                                          trend = TRUE,
                                          FDR = 0.05)
  
  #filter only significant KEGG terms
  kegga_treat_sig[[x]] <- topKEGG(results = kegga_treat[[x]], number = 30)
} 

# save rds object for use in downstream GO and KEGG analysis
kegga_treat %>% saveRDS(file = "../0_data/rds_object/kegga_treat_unfiltered.rds")
kegga_treat_sig %>% saveRDS(file = "../0_data/rds_object/kegga_treat_filtered.rds")

#save each unfiltered comparison group in the output directory
writexl::write_xlsx(x = kegga_treat,
                    path = "../3_output/kegga_treat_unfiltered.xlsx")

```

<!-- #### QLF -->

```{r qlfKegga, eval=FALSE}
# Create list object
kegga_qlf=list()
kegga_qlf_sig=list()
for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  #populate the kegga list with all KEGG terms
  kegga_qlf[[x]] <- edgeR::kegga.DGELRT(de = qlf[[x]], 
                                          geneid = qlf[[x]]$genes$entrezid, 
                                          species = "Hs",
                                          trend = TRUE,
                                          FDR = 0.05)
  
  #filter only significant KEGG terms
  kegga_qlf_sig[[x]] <- topKEGG(results = kegga_qlf[[x]], number = 30)
} 

# save rds object for use in downstream GO and KEGG analysis
kegga_qlf %>% saveRDS(file = "../0_data/rds_object/kegga_qlf_unfiltered.rds")
kegga_qlf_sig %>% saveRDS(file = "../0_data/rds_object/kegga_qlf_filtered.rds")

#save each unfiltered comparison group in the output directory
writexl::write_xlsx(x = kegga_qlf,
                    path = "../3_output/kegga_qlf_unfiltered.xlsx")
```

# Visualisation

## Heatmap

```{r coolmap, eval=FALSE, include=FALSE}
logCPM <- cpm(dge, prior.count=2, log=TRUE)
rownames(logCPM) <- dge$genes$gene
colnames(logCPM) <- paste(dge$samples$sample_group, 1:3, sep="-")

o <- order(treat$CKIvsOMT$table$PValue)
logCPM <- logCPM[o[1:30],]

coolmap(x = p53_treat_UTvsCKI_test, 
        margins=c(7,7), lhei=c(1,5), lwid=c(1,3))
```


```{r}
kegg_id <- "hsa04657"
kegg_pathway <- KEGGREST::keggGet(kegg_id)
kegg_pathway_name <- kegg_pathway[[1]]$NAME

entrezID=list()
matrix=list()

for (i in 1:ncol(contrast)){
  #at each iteration, let x = name of each contrast group
  x=comparison_group[i,]
  
  entrezID[[x]] <- enrichKEGG_treat_sig[[x]][,c("ID", "geneID")] %>% as.data.frame()
  entrezID[[x]] <- entrezID[[x]][kegg_id,"geneID"] %>% as.data.frame()
  entrezID[[x]] <- separate_rows(entrezID[[x]],.,sep = "/")
  colnames(entrezID[[x]]) <- "entrezid"
  
  match <- treat_filtered[[x]]$entrezid %in% entrezID[[x]]$entrezid

  matrix[[x]] <- treat_filtered[[x]][match, c("gene_name", "logFC")]
}

matrix_merged <- matrix$`UT vs CKI` %>% 
  left_join(matrix$`CKI-Mac vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Nme vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Omt vs CKI`, by = "gene_name") %>% 
  left_join(matrix$`CKI-Tri vs CKI`, by = "gene_name") %>% 
  column_to_rownames("gene_name")

colnames(matrix_merged) <- c("UT\nvs\nCKI", "CKI-Mac\nvs\nCKI", "CKI-Nme\nvs\nCKI", "CKI-Omt\nvs\nCKI", "CKI-Tri\nvs\nCKI")



my_palette <- colorRampPalette(c("dodgerblue3", "white", "firebrick3"))(n = 201)

pheatmap(mat = matrix_merged,
         clustering_distance_rows = "euclidean",
         treeheight_row = 100,
         legend_labels = "logFC",
         main = paste0("Heatmap of ", kegg_pathway_name),
         fontsize = 16,
         fontsize_row = 14,
         fontsize_number = 10,
         fontsize_col = 14,
         angle_col = 0,
         border_color = "gray60",
         color = my_palette)



```



# Setup
```{r sessionInfo}
sessionInfo()

```

